---
title: "HPD.explore"
output: html_document
author: Max Gordy
date: "2025-03-12 (update 03-19)"
---

```{r setup, include=FALSE}
library(tidycensus)
library(tidyverse)
library(sf)
library(ggplot2)
library(janitor)
library(readxl)
library(gt)
library(ggpattern)
library(leaflet)
library(mapview)
library(dplyr)
library(leafpop)
```

# HPD Affordable Housing Production Data

HPD reports on projects, buildings, and units that began after January 1, 2014, and are counted towards either the Housing New York plan (1/1/2014 – 12/31/2021) or the Housing Our Neighbors: A Blueprint for Housing & Homelessness plan (1/1/2022 – present).
Complete info on data is available [here](https://data.cityofnewyork.us/Housing-Development/Affordable-Housing-Production-by-Project/hq68-rnsi/about_data).

### General Notes

- Available by project and by building \ 
  - building has bbls \
  - project has program info \
  - I'm joining project info to individual buildings \
- 8,108 building entries
  1604 marked confidential - no location data
- 3037 entries that overlap with our filtered DCP dataset \
	- 2147 unique overlapping projects \
		- ex: [New Penn Development](https://architizer.com/projects/new-penn-development/) is a complex of 10 affordable housing buildings, so it has 10 entries in this dataset. \
- Filtering by new construction only \ 
  - (3,946/8,108)
    - 970/3,946 are marked confidential

## Data

```{r, include=FALSE}
# load in data

hpd_projs <- read_csv("data/hpd/Affordable_Housing_Production_by_Project_20250312.csv") %>% clean_names()

head(hpd_projs)


hpd_bldgs <- read_csv("data/hpd/Affordable_Housing_Production_by_Building_20250312.csv") %>% 
  clean_names() %>% 
  mutate(bbl = as.character(bbl)) %>% 
  filter(
    reporting_construction_type == "New Construction"
  )

head(hpd_bldgs)

```

```{r NA counts, echo=FALSE}
print("Column Names:")
colnames(hpd_bldgs)

print("Total Entries:")
length(hpd_bldgs$`project_name`)

print("Entries Marked Confidential:")
sum(hpd_bldgs$`project_name` == "CONFIDENTIAL")

```


## Mapping

You can click on points to bring up bbl and project name. 

```{r echo=FALSE}
# mapping
hpd_sf <- hpd_bldgs %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  mutate(unit_type = case_when(
    !is.na(counted_homeownership_units) & counted_homeownership_units > 0 & is.na(counted_rental_units) ~ "Homeownership",
    !is.na(counted_rental_units) & counted_rental_units > 0 & is.na(counted_homeownership_units) ~ "Rental",
    TRUE ~ "Mixed"
  )) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Create popup table with project name and BBL
hpd_sf$popup_info <- data.frame(
  "Project Name" = hpd_sf$project_name,
  "BBL" = hpd_sf$bbl,
  "Total Units" = hpd_sf$all_counted_units
)

# Create interactive map with size = total units, color = unit type, and popups showing project name and BBL
mapview(hpd_sf, 
        zcol = "unit_type", 
        cex = "all_counted_units", 
        col.regions = c("blue", "purple", "red"),
        popup = popupTable(hpd_sf, zcol = c("project_name", "bbl", "all_counted_units")))

```


## Program Analysis

It seems like the number of confidential program groups and NA tax benefits makes this dataset not very useful for program analysis.

```{r include=FALSE}

# Get the column names of both data frames
cols_bldgs <- colnames(hpd_bldgs)
cols_projs <- colnames(hpd_projs)

# Find columns in df1 that are not in df2
unique_cols_in_projs <- setdiff(cols_projs, cols_bldgs)

# Print the result
print(paste("Variables in Projects but not Buildings:", paste(unique_cols_in_projs, collapse = ", ")))

# Select only the unique columns in projs
projs_unique <- hpd_projs %>% select(project_id, program_group, planned_tax_benefit, senior_units)

# Join projs to buildings on a common key (e.g., project_id or bbl)
joined_data <- hpd_bldgs %>%
  left_join(projs_unique, by = "project_id")  # Replace 'project_id' with your actual key

```

### Units by Program Group
Removed Confidential Entries (3,823,936/4,097,311)

```{r, echo=FALSE}

program_group_summary <- joined_data %>%
  mutate(program_group = ifelse(is.na(program_group), "NA", program_group)) %>%
  group_by(program_group) %>%
  summarise(
    count = n(),
    total_units = sum(all_counted_units, na.rm = TRUE)
  ) %>%
  arrange(desc(count))

# Print the table
print(program_group_summary)

# Create the pie chart
ggplot(program_group_summary, aes(x = "", y = count, fill = program_group)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +  # Removes background grid and axes
  labs(fill = "Program Group", title = "Distribution of Program Group Designations") +
  theme(legend.position = "right")





# Summarize rental and homeownership units by program group
summary_data <- joined_data %>%
  filter(program_group != "CONFIDENTIAL") %>%  # Exclude confidential program groups if needed
  group_by(program_group) %>%  # Group by program group
  summarise(
    rental_units = sum(counted_rental_units, na.rm = TRUE),
    homeownership_units = sum(counted_homeownership_units, na.rm = TRUE),
    total_units = sum(all_counted_units, na.rm = TRUE)  # Add total units for sorting
  )

# Reshape data to long format for easier plotting
summary_data_long <- summary_data %>%
  pivot_longer(cols = c(rental_units, homeownership_units),  # Exclude 'total_units'
               names_to = "unit_type", values_to = "unit_count") %>%
  # Reorder 'program_group' based on total 'unit_count'
  mutate(program_group = reorder(program_group, unit_count, sum))

# Plot the data with stacked rental and homeownership units
ggplot(summary_data_long, aes(x = program_group, y = unit_count, fill = unit_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Rental and Homeownership Units by Program Group",
    x = "Program Group",
    y = "Number of Units",
    fill = "Unit Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Slant x-axis labels
  )

```


### Units by Tax Benefit
NAs removed (2,570,097 NAs out of 2,579,320)
```{r, echo=FALSE}

tax_benefit_summary <- joined_data %>%
  mutate(planned_tax_benefit = ifelse(is.na(planned_tax_benefit), "NA", planned_tax_benefit)) %>%
  group_by(planned_tax_benefit) %>%
  summarise(
    count = n(),
    total_units = sum(all_counted_units, na.rm = TRUE)
  ) %>%
  arrange(desc(count))

# Print the table
print(tax_benefit_summary)

# Create the pie chart
ggplot(tax_benefit_summary, aes(x = "", y = count, fill = planned_tax_benefit)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +  # Removes background grid and axes
  labs(fill = "Planned Tax Benefit", title = "Distribution of Planned Tax Benefit Designations") +
  theme(legend.position = "right")

# Summarize rental and homeownership units by program, excluding NAs
summary_data <- joined_data %>%
  filter(!is.na(planned_tax_benefit)) %>%  # Exclude rows where 'planned_tax_benefit' is NA
  group_by(planned_tax_benefit) %>%
  summarise(
    rental_units = sum(counted_rental_units, na.rm = TRUE),
    homeownership_units = sum(counted_homeownership_units, na.rm = TRUE),
    total_units = rental_units + homeownership_units  # Calculate total units
  )

# Reorder 'planned_tax_benefit' based on total_units in descending order
summary_data <- summary_data %>%
  mutate(planned_tax_benefit = reorder(planned_tax_benefit, total_units, sum))

# Reshape data to long format for easier plotting
summary_data_long <- summary_data %>%
  pivot_longer(cols = c(rental_units, homeownership_units), 
               names_to = "unit_type", values_to = "unit_count")

# Plot the data with stacked rental and homeownership units
ggplot(summary_data_long, aes(x = planned_tax_benefit, y = unit_count, fill = unit_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Rental and Homeownership Units by Tax Benefit",
    x = "Planned Tax Benefit",
    y = "Number of Units",
    fill = "Unit Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Slant x-axis labels
  )


```


### Units by Project 

Confidential Removed, vast majority of units (3,823,936/4,097,311)
```{r, echo=FALSE}

project_name_summary <- joined_data %>%
  mutate(project_name = ifelse(is.na(project_name), "NA", project_name)) %>%
  group_by(project_name) %>%
  summarise(
    count = n(),
    total_units = sum(all_counted_units, na.rm = TRUE)
  ) %>%
  arrange(desc(count))

# Print the table
print(project_name_summary)


# Summarize rental and homeownership units by program
summary_data <- joined_data %>%
  filter(project_name != "CONFIDENTIAL") %>%  # Exclude "CONFIDENTIAL" projects
  group_by(project_name) %>%  # Group by project name
  summarise(
    rental_units = sum(counted_rental_units, na.rm = TRUE),
    homeownership_units = sum(counted_homeownership_units, na.rm = TRUE),
    total_units = sum(all_counted_units, na.rm = TRUE)
  )

# Reshape data to long format for easier plotting, excluding 'total_units'
summary_data_long <- summary_data %>%
  top_n(10, total_units) %>%  # Select top 10 programs based on total units
  pivot_longer(cols = c(rental_units, homeownership_units),  # Exclude total_units
               names_to = "unit_type", values_to = "unit_count") %>%
  # Reorder 'project_name' based on total 'unit_count'
  mutate(project_name = reorder(project_name, unit_count, sum))

# Plot the data with stacked rental and homeownership units
ggplot(summary_data_long, aes(x = project_name, y = unit_count, fill = unit_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Rental and Homeownership Units by Project",
    x = "Project Name",
    y = "Number of Units",
    fill = "Unit Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Slant x-axis labels
  )

```

# Analyze by PUMA

# pull data
```{r}
# Load 2020 PUMA shapefile
nyc_pumas <- st_read("data/nyc_mappluto_24v4_fgdb/nypuma2020_25a/nypuma2020.shp") %>%
  st_transform(crs = st_crs(hpd_sf))  # Match CRS with hpd_sf
```

### Spatial Join and Summarize
```{r}
# Only use geocoded HPD buildings
hpd_with_puma <- hpd_sf %>%
  filter(!is.na(all_counted_units)) %>%
  st_join(nyc_pumas, join = st_within)  # Join buildings to PUMAs

hpd_puma_summary <- hpd_with_puma %>%
  st_drop_geometry() %>%  # Drop geometry to enable summarization
  group_by(PUMA) %>%  # Adjust this to match actual column name in your shapefile, maybe `puma` or `puma_code`
  summarise(total_units = sum(all_counted_units, na.rm = TRUE)) %>%
  arrange(desc(total_units))

```

#### join back and map
```{r}
puma_with_hpd <- nyc_pumas %>%
  left_join(hpd_puma_summary, by = "PUMA")  # Adjust field name if needed

# Quick choropleth map
ggplot(puma_with_hpd) +
  geom_sf(aes(fill = total_units)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey80") +
  theme_minimal() +
  labs(
    title = "Affordable Housing Units (New Construction) by PUMA",
    fill = "Total Units"
  )

leaflet(puma_with_hpd) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    fillColor = ~colorNumeric("YlOrRd", domain = total_units)(total_units),
    weight = 1,
    color = "#333",
    fillOpacity = 0.7,
    popup = ~paste("PUMA:", PUMA, "<br>Total Units:", total_units)
  )

```


